Memes: an R interface to the MEME Suite

### 2020-11-11 manuscript.Rmd
Fix figure callouts.

---
title: 'Memes: an R interface to the MEME Suite'
author:
- ORCID: 0000-0003-1000-1579
  name: Spencer L. Nystrom
- ORCID: 0000-0001-8226-0604
  name: Daniel J. McKay
output: 
 - word_document
---

```{r setup, echo = FALSE, warning=FALSE, message=FALSE}
# Figure caption: https://stackoverflow.com/questions/31926623/figures-captions-and-labels-in-knitr
suppressPackageStartupMessages(library(memes))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(GenomicRanges))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ComplexHeatmap))
drake::r_make()
```

@@

These data are readily visualized using the `universalmotif::view_motifs`
function. Visualization of the results reveals two distinct motifs associated
with decreasing and increasing sites \ref{fig:dec_v_inc_dreme}. In this
analysis, the *de-novo* discovered motifs within each category appear visually
similar to each other, however, the MEME Suite does not provide a mechanism to
compare groups of motifs based on all pairwise similarity metrics. We utilized
the `universalmotif::compare_motifs` function to compute Pearson correlation
coefficients for each set of motifs to quantitatively assess motif similarity
\ref{fig:dec_v_inc_dreme}.

```{r dec_v_inc_dreme, fig.height=10, fig.width=10, echo=F, fig.cap="\\label{fig:dec_v_inc_dreme}PWMs of discovered motifs in Decreasing (A) and Increasing (B) sites. C,D Pearson correlation heatmaps comparing similarity of Decreasing (C) and Increasing (D) motifs."}
patchwork::wrap_plots(dreme_fig_top, dreme_fig_bottom, ncol = 1) +
  patchwork::plot_annotation(tag_levels = "A")
```

@@

Using this approach, the results can be visualized using  the
`view_tomtom_hits()` function to visually inspect the matches assigned by
TomTom, providing a simple way to assess the quality of the match.  A
representative plot of these data is shown in \ref{fig:tomtom_example}, while
the full results can be viewed in \ref{tab:tomtom_table}. This analysis revealed
that the motifs associated with decreasing sites are highly similar to the E93
motif. Conversely, motifs found in increasing sites match the transcription
factors br, l(3)neo38, gl, and lola \ref{tab:tomtom_table}. These data support
the hypothesis that E93 binding to its motif may play a larger role in chromatin
closing activity, while binding of the transcription factor br is associated
with chromatin opening.

```{r tomtom_example, echo = F, fig.height=3, fig.width=7.5, fig.cap="\\label{fig:tomtom_example}Representative plots generated by view_tomtom_hits showing the top hit for motifs discovered in decreasing (A) and increasing (B) peaks."}
plot_top_hit <- function(x){
  x %>% 
    view_tomtom_hits(1)
}

@@

dreme_vs_static_tomtom %>% 
  purrr::imap(~{
    .x %>% 
      dplyr::mutate(name = paste0(.y, " ", name)) %>% 
      update_motifs() %>% 
      plot_top_hit() %>% 
      .[[2]]
    }) %>% 
  patchwork::wrap_plots() +
    patchwork::plot_annotation(tag_levels = "A")
```
```{r, echo=F}
library(gt)

dreme_vs_static_tomtom %>% 
  dplyr::bind_rows(.id = "category") %>% 
  dplyr::mutate(category = paste(category, "Motifs")) %>% 
  dplyr::arrange(category, rank) %>% 
  dplyr::select(category, name, best_match_altname) %>% 
  dplyr::rename("Best Match TF" = best_match_altname) %>% 
  gt(rowname_col = "name",
     groupname_col = "category")
     # this feature is in development branch??
     # search SO: gt table reference
     #lab = "tab:tomtom_table")

@@

`runAme` syntax is identical to `runDreme` in that discriminative designs can be
constructed by calling list entries by name. Data can be visualized using
`ame_plot_heatmap()`, revealing that the E93 motif is strongly enriched at
Decreasing sites, while l(3)neo38, lola, and br are enriched in Increasing
sites, supporting the *de-novo* discovery results \ref{fig:ame_heatmap}.
Additionally, AME detects several other transcription factor motifs that
distinguish decreasing and increasing sites, providing additional clues to
potential factors that also bind with E93 affect chromatin accessibility.

@@

```{r ame_heatmap, fig.height=3, fig.width=7.5, echo = FALSE, fig.cap="\\label{fig:ame_heatmap} A heatmap of the -log10(adj.pvalues) from AME enrichment testing within increasing and decreasing sites."}
ame_vs_static %>% 
  dplyr::bind_rows(.id = "category") %>% 
  dplyr::mutate(category = paste0(category, "\nAccessibility")) %>% 
  dplyr::group_by(category, motif_alt_id) %>% 
  dplyr::filter(adj.pvalue == min(adj.pvalue)) %>% 
  ame_plot_heatmap(group = category, id = motif_alt_id) +
    labs(x = "Transcription Factor",
         y = "ChIP Peak Type")
```

@@

Using this approach we can deeply examine the properties of the E93 motif within
each chromatin response category. First, by counting the number of E93 motifs
within each category, we demonstrate that Decreasing sites are more likely than
increasing or static sites to contain an E93 motif \ref{fig:fimo}A. We extend
these observations by rederiving position-weight matrices from sequences
matching the E93 motif within each category, allowing visual inspection of motif
quality across groups \ref{fig:fimo}B. Notably, differences in quality at base
positions 8-12 appear to distinguish increasing from decreasing motifs, where
decreasing motifs are more likely to have strong A bases at positions 8 and 9,
while E93 motifs from increasing sites are more likely to have a T base pair at
position 12 \ref{fig:fimo}B. Examination of bulk FIMO scores (where higher
scores represent motifs more similar to the reference) also reveals differences
in E93 motif quality between groups, in particular, E93 motifs from Decreasing
sites have higher scores \ref{fig:fimo}C. Together, these data demonstrate that
a key distinguishing factor between whether a sites will decrease or increase in
chromatin accessibility following E93 binding is the number and quality of E93
motifs at that site.

@@

```{r fimo, fig.width=7.5, fig.height=9, fig.cap="\\label{fig:fimo}A. Stacked barplot showing fraction of each chromatin response category containing E93 motif matches. B. PWMs generated from E93 motif sequences detected in each chromatin response category. C. Boxplot of FIMO Score for each E93 motif within each chromatin response category. Outliers are plotted as distinct points."}
#patchwork::wrap_plots(n_motifs, motif_quality, patchwork::guide_area(), e93_stack, widths = c(0.2, 0.3, 0.5))
patchwork::wrap_plots(n_motifs, patchwork::guide_area(), 
                      e93_stack, motif_quality, 
                      ncol = 2, widths = c(0.6, 0.4),
                      guides = 'collect') +
    patchwork::plot_annotation(tag_levels = "A")
```

@@

This manuscript was automatically generated using Rmarkdown within the
`memes_docker` container (TODO: ADD VERSION). Its source code, raw data, and
instructions to reproduce all analysis can be found at
[github.com/snystrom/memes_paper/](github.com/snystrom/memes_paper/).

>>>>

---
title: 'Memes: an R interface to the MEME Suite'
author:
- ORCID: 0000-0003-1000-1579
  name: Spencer L. Nystrom
- ORCID: 0000-0001-8226-0604
  name: Daniel J. McKay
output: 
  bookdown::word_document2:
    fig_caption: yes
---

```{r setup, echo = FALSE, warning=FALSE, message=FALSE, include=FALSE}
# Figure caption: https://stackoverflow.com/questions/31926623/figures-captions-and-labels-in-knitr
suppressPackageStartupMessages(library(memes))
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(GenomicRanges))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ComplexHeatmap))
drake::r_make()
```

@@

These data are readily visualized using the `universalmotif::view_motifs`
function. Visualization of the results reveals two distinct motifs associated
with decreasing and increasing sites (**Figure \@ref(fig:dec-v-inc-dreme)**). In this
analysis, the *de-novo* discovered motifs within each category appear visually
similar to each other, however, the MEME Suite does not provide a mechanism to
compare groups of motifs based on all pairwise similarity metrics. We utilized
the `universalmotif::compare_motifs` function to compute Pearson correlation
coefficients for each set of motifs to quantitatively assess motif similarity
(**Figure \@ref(fig:dec-v-inc-dreme)**).

```{r dec-v-inc-dreme, fig.height=10, fig.width=10, echo=F, fig.cap="PWMs of discovered motifs in Decreasing (A) and Increasing (B) sites. C,D Pearson correlation heatmaps comparing similarity of Decreasing (C) and Increasing (D) motifs."}
patchwork::wrap_plots(dreme_fig_top, dreme_fig_bottom, ncol = 1) +
  patchwork::plot_annotation(tag_levels = "A")
```

@@

Using this approach, the results can be visualized using  the
`view_tomtom_hits()` function to visually inspect the matches assigned by
TomTom, providing a simple way to assess the quality of the match.  A
representative plot of these data is shown in **Figure \@ref(fig:tomtom-example)**, while
the full results can be viewed in **Table \@ref(tab:tomtom-table)**. This analysis revealed
that the motifs associated with decreasing sites are highly similar to the E93
motif. Conversely, motifs found in increasing sites match the transcription
factors br, l(3)neo38, gl, and lola (**Table \@ref(tab:tomtom-table)**). These data support
the hypothesis that E93 binding to its motif may play a larger role in chromatin
closing activity, while binding of the transcription factor br is associated
with chromatin opening.

```{r tomtom-example, echo = F, fig.height=3, fig.width=7.5, fig.cap="Representative plots generated by view_tomtom_hits showing the top hit for motifs discovered in decreasing (A) and increasing (B) peaks."}
plot_top_hit <- function(x){
  x %>% 
    view_tomtom_hits(1)
}

@@

dreme_vs_static_tomtom %>% 
  purrr::imap(~{
    .x %>% 
      dplyr::mutate(name = paste0(.y, " ", name)) %>% 
      update_motifs() %>% 
      plot_top_hit() %>% 
      .[[2]]
    }) %>% 
  patchwork::wrap_plots() +
    patchwork::plot_annotation(tag_levels = "A")
```

```{r, echo=F}
library(gt)

dreme_vs_static_tomtom %>% 
  dplyr::bind_rows(.id = "category") %>% 
  dplyr::mutate(category = paste(category, "Motifs")) %>% 
  dplyr::arrange(category, rank) %>% 
  dplyr::select(category, name, best_match_altname) %>% 
  dplyr::rename("Best Match TF" = best_match_altname) %>% 
  gt(rowname_col = "name",
     groupname_col = "category") %>% 
  tab_header(title = "Best match results from TomTom")
     # this feature is in development branch??
     # search: R markdown gt table reference
     # https://community.rstudio.com/t/how-to-set-a-reference-in-rmarkdown-to-a-gt-table/23781/2
     #lab = "tab:tomtom-table")

@@

`runAme` syntax is identical to `runDreme` in that discriminative designs can be
constructed by calling list entries by name. Data can be visualized using
`ame_plot_heatmap()`, revealing that the E93 motif is strongly enriched at
Decreasing sites, while l(3)neo38, lola, and br are enriched in Increasing
sites, supporting the *de-novo* discovery results (**Figure \@ref(fig:ame-heatmap)**).
Additionally, AME detects several other transcription factor motifs that
distinguish decreasing and increasing sites, providing additional clues to
potential factors that also bind with E93 affect chromatin accessibility.

@@

```{r ame-heatmap, fig.height=3, fig.width=7.5, echo = FALSE, fig.cap="A heatmap of the -log10(adj.pvalues) from AME enrichment testing within increasing and decreasing sites."}
ame_vs_static %>% 
  dplyr::bind_rows(.id = "category") %>% 
  dplyr::mutate(category = paste0(category, "\nAccessibility")) %>% 
  dplyr::group_by(category, motif_alt_id) %>% 
  dplyr::filter(adj.pvalue == min(adj.pvalue)) %>% 
  ame_plot_heatmap(group = category, id = motif_alt_id) +
    labs(x = "Transcription Factor",
         y = "ChIP Peak Type")
```

@@

Using this approach we can deeply examine the properties of the E93 motif within
each chromatin response category. First, by counting the number of E93 motifs
within each category, we demonstrate that Decreasing sites are more likely than
increasing or static sites to contain an E93 motif (**Figure \@ref(fig:fimo)A**). We extend
these observations by rederiving position-weight matrices from sequences
matching the E93 motif within each category, allowing visual inspection of motif
quality across groups (**Figure \@ref(fig:fimo)B**). Notably, differences in quality at base
positions 8-12 appear to distinguish increasing from decreasing motifs, where
decreasing motifs are more likely to have strong A bases at positions 8 and 9,
while E93 motifs from increasing sites are more likely to have a T base pair at
position 12 (**Figure \@ref(fig:fimo)B**). Examination of bulk FIMO scores (where higher
scores represent motifs more similar to the reference) also reveals differences
in E93 motif quality between groups, in particular, E93 motifs from Decreasing
sites have higher scores (**Figure \@ref(fig:fimo)C**). Together, these data demonstrate that
a key distinguishing factor between whether a sites will decrease or increase in
chromatin accessibility following E93 binding is the number and quality of E93
motifs at that site.

@@

```{r fimo, echo = F, fig.width=7.5, fig.height=9, fig.cap="A. Stacked barplot showing fraction of each chromatin response category containing E93 motif matches. B. PWMs generated from E93 motif sequences detected in each chromatin response category. C. Boxplot of FIMO Score for each E93 motif within each chromatin response category. Outliers are plotted as distinct points."}
#patchwork::wrap_plots(n_motifs, motif_quality, patchwork::guide_area(), e93_stack, widths = c(0.2, 0.3, 0.5))
patchwork::wrap_plots(n_motifs, patchwork::guide_area(), 
                      e93_stack, motif_quality, 
                      ncol = 2, widths = c(0.6, 0.4),
                      guides = 'collect') +
    patchwork::plot_annotation(tag_levels = "A")
```

@@

This manuscript was automatically generated using Rmarkdown within the
`memes_docker` container (**TODO: ADD VERSION**). Its source code, raw data, and
instructions to reproduce all analysis can be found at
[github.com/snystrom/memes_paper/](github.com/snystrom/memes_paper/).
